Haskell in Computational Biology

### 2013-05-28 jfp.tex

\newcommand\figref[1]{Figure~\ref{fig:#1}}
\newcommand\figrefpair[3][and]{Figures \ref{fig:#2}~#1~\ref{fig:#3}}
\newcommand\figlabel[1]{\label{fig:#1}}
\newcommand\secref[1]{Section~\ref{sec:#1}}
\newcommand\seclabel[1]{\label{sec:#1}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\item
\emph{[POTENTIAL BULLETS ABOUT SPEED AND ``MORE FUNCTIONAL'']}
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds to the original paper by describing our experience in
trying to simplify and speed up the central algorithm.
AND FIX THE CODE AND...

@@

A~full hidden Markov model also has distinguished ``begin'' and ``end''
states; begin and end states are like match states, but they
correspond to no amino acids and contain no emission probabilities.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, 
the begin and end states are known by their positions:
in a $\alignwidth$-column model, the begin state is the only state in
column~0 and the end state is the only state in column~$\alignwidth$.
Every other state is referred to by its column~$j$
and by one of these labels:
\smallverbatiminput{statelabel}
We~group states that share a column~$j$; these states are referred
to collectively as ``node~$j$,'' as shown in \figref{plan7}.
This model
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.
\remark{We need to figure out if the begin state is really node 0}

@@

All scores computed by \mrfy\ are sums of these transition and
emission scores.
Because these sums can be usefully attached to many types of values,
we~have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \mono{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
\remark{Lazy in the first arg is about 1\%~slower}
The \texttt{/+/} function adds to the score without touching the
 container.
Function \texttt{fmap} is also defined; it~applies a function to a
 container's contents. 
%
Finally, we~made \texttt{Scored} an instance of~\texttt{Ord}.
Containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

@@

\begin{figure}%%%[h!]
\centerline{\includegraphics{speedup}}
\medskip
\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
\figlabel{speedup}
\end{figure}

@@

@@

The code example for \verb+vee'+ on page~\pageref{code:vee-prime}
shows the computation of the general case of~$V^{\prime M}_j(i)$ from 
\figref{viterbi-transformed}.
But the full story is not nearly as nice at one could hope for:
in~the code that accompanies our 2012 ICFP paper,
\verb+vee'+ has 25~cases overall.
Many of these cases exist because we~did not have enough insight into
Haskell, and we made some decisions that might have been more
appropriate to a FORTRAN program:
\begin{itemize}
\item
Recursive computations were done on smaller array indices, not smaller
models.
\item
Array indices 1, 0, and $-1$ were all used to indicate special
conditions.
\item
Senseless combinations of array indices and states were handled by
assigning each such combination the largest representable score, which
corresponds to the smallest representable probability.
\end{itemize}
Before trying to speed up the code, we~simplified~it.

Our first step in simplification was to start thinking like Haskell
programmers and use types:
\begin{itemize}
\item
We~refactored the single, flat \texttt{Node} type from our ICFP paper
into the three state types you see on page~\pageref{code:model3-node}.
\item 
We introduced a \texttt{Model} type that consists of a special Begin
node containing a begin and an insertion state, an~unbounded sequence
of ordinary \texttt{Nodes}, and an end state.
And we represented the sequence as a list, not an array.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only match, insertion, and deletion states, so we removed the
\texttt{StateLabel} value constructors corresponding to begin and end
states, leaving only the three constructors shown on
page~\ref{code:state-label}. 
(We~moved the begin and end labels to another type, which is used only
to prettyprint models.) 
\end{itemize}
The immediate effect of these changes was to reduce case analysis on
two dimensions:
instead of five state labels, we got three;
and instead of three special array indices, we got just one special
list (the empty list).

@@

The version of \verb+vee'+ shown in  \figref{viterbi-hov4} is abstract
in another way: the type of the result is the unknown type~\texttt{a}.
Type~\texttt{a} represents

\figref{viterbi-abstracted} doesn't deal with the begin state.
In~our legacy file format, and also in our code, the begin state is
represented as a match state inside an ordinary node.
In~a perfect world, then, we would add one case to say that the match
state with $i=j=0$ is reachable at zero cost, and we'd be finished.
But the world is imperfect, because although the begin state is
represented as a match state, unlike a match state, the begin state
does not emit.
We~therefore need a special case for each of the three states that can
follow the begin state.

Here is the code:
NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
NEED TO EXPLAIN [[transition]] and [[emission]]

As a result of these changes, in the revised code,
\begin{itemize}
\item
Only array index~0 is used to indicate a special
condition: the transition from the leftmost middle node to the unique
begin node.
\item
\end{itemize}

\begin{verbatim}
### 2013-05-27 01:23   Running med(vpasses=1): icfp-branch mrfy nr-jfp
icfp-branch faab8dc       54.07s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
mrfy d5604bc*             44.90s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
nr-jfp 00ef54e*           47.59s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
\end{verbatim}

\smallverbatiminput{hov-prevs}

The main benefit is that the Viterbi function is reduced from 90~lines
of code to just 38~lines, of which 8~lines are type signatures.
This change laid the groundwork for more experiments.

If~we look not at the whole package but just at Viterbi's algorithm,
the~performance story is even more grim:
we~wrote a custom C~program that computes a score using Viterbi's
algorithm and does nothing else, and it runs about a hundred times
faster than \mrfy's implementation.

BENCHMARK RESULTS WERE WILDLY INCONSISTENT.

THE STORY OF THE UNPACK PRAGMAS; CITING JOHAN TIBELL

\subsection{Awkward debugging and testing}

@@

Our experience writing code and trying new ideas was excellent, as~was
the ease of parallelizing \mrfy.
Higher-order functions, memoization, laziness, and parallel strategies
really worked.
But~we also encountered obstacles that prevented functional
programming from working as well as we would have liked.
The~most significant obstacles were in debugging and testing.

We had a hard time diagnosing run-time errors.
We~expected some run-time errors;
our~group's legacy file
format is poorly documented and hard to deal with.
(When beta
strands overlap and are doubly paired, even the invariants of the
format are unclear.)
But using Haskell, we~found the errors hard to diagnose.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~didn't know about the backtrace feature of GHC's profiler,
and even after we learned about it, it didn't help:
the profiler can be used only on very small test cases,
which didn't always trigger the errors.
This same limitation affected GHCi's debugger; 
in~GHCi, our \texttt{vee'} function
is too slow to be runnable on nontrivial
inputs.
Moreover,  GHCi's debugger can set breakpoints only at top-level functions
or at specific column/line positions, which made debugging the memoized
\texttt{vee''} function impractical.
In~August~2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

@@

Only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles that prevented the
junior people from using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
predominantly on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
(For programmers accustomed to object-oriented dispatch
on \emph{arguments}, it is hard to grasp how Haskell's type-class
system can find the proper version
of \texttt{arbitrary} using only the type of a \emph{value}.)
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did 
use QuickCheck, the counterexamples were too large to be
informative. 
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only when the whole team pitched in to
meet the deadlines for this paper.

At the last minute, QuickCheck did find
a bug in our implementation of Viterbi's algorithm: we had omitted
the score for the transition from the final node of the hidden
Markov model to the special ``end'' state.
Without QuickCheck, we~probably wouldn't have known anything was wrong.
%it is unlikely that we would have found this bug,
%and if we had, determining the cause would have required much time.

@@

If~you are a computational biologist and you are interested in
functional programming, you don't need extensive preparation
to be productive in Haskell.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell 
for medium-sized projects,
but his contributions to MRFy have been limited, mostly to \emph{post hoc}
refactoring and testing.

@@

We~had quite some difficulty profiling, but we hope that
this difficulty may be mitigated by new profiling tools released early
in~2012 with
GHC~7.4.
GHC assigns costs to ``cost centers'' \cite{sansom-pj},
and
in GHC~7.0, which we used 
for most of MRFy's development, 
cost-center annotations had to be added manually
to nested functions.
Although these annotations
made our code so ugly that we felt compelled to
remove them, they did enable us to improve the performance of~\verb+vee'+ as
discussed in \secref{perf}.
%We~miss valgrind, whose profiler gives useful information
%without \emph{any} annotations.
GHC~7.4 provides 
more sophisticated profiling tools, which we look forward to using.
Difficulties using Cabal to enable profiling of installed libraries
may remain.

@@

Ideally we would use QuickCheck to find bugs, but as we mention in
\secref{awkward-quickcheck}, we found obstacles.
We~have now overcome these obstacles, but
we~sorely regret not doing so earlier.
In~light of our experience, we will institute a new programming practice:
whenever we introduce a new data type,
we~will write the instance of
\texttt{Arbitrary} right away, while
relevant invariants are still fresh in memory.
When invariants are not enforced by Haskell's static type system,
we~will write them as Haskell predicates.
We can then immediately run QuickCheck on each predicate, to verify
that our Arbitrary instance agrees with the predicate.
For each predicate~\texttt{p} we can also check
\mono{fmap (all p . shrink) arbitrary}.

@@

If~you want to use Haskell in your research, we~believe that you must
have enough experience with functional programming that you can build
\emph{all} the code you need, not only the code that is easy to write
in a functional language.
Implementing Viterbi's equations in Haskell was pure joy.
Writing an iterative search in purely functional style was~easy.
Transforming data in the HMMER file format, \emph{without}
using mutable state the way the C++ code does, was difficult.

NEW PARAGRAPH: STEP SLOWLY AWAY FROM YOUR CODE, AND SEE IF YOU THINK
IT RESEMBLES ANYTHING A HASKELL PROGRAMMER WOULD HAVE WRITTEN.

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
John Tibell taught us the \texttt{UNPACK} pragma.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.

@@

\section{THIS SECTION WILL BE DELETED.  SAUVE QUI PEUT!}

\input{jfpnotes}

>>>>

\newcommand\figref[1]{Figure~\ref{fig:#1}}
\newcommand\figrefpair[3][and]{Figures \ref{fig:#2}~#1~\ref{fig:#3}}
\newcommand\figlabel[1]{\label{fig:#1}}
\newcommand\secref[1]{Section~\ref{sec:#1}}
\newcommand\secrefpair[3][and]{Sections \ref{sec:#2}~#1~\ref{sec:#3}}
\newcommand\seclabel[1]{\label{sec:#1}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secrefpair{improving-viterbi}{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds a new technical section (\secref{improving-viterbi})
which describes our experience 
refactoring and improving our implementation of Viterbi's 
\citeyearpar{Viterbi:1967hq}
algorithm.
We~have also simplified the stochastic search as well as other elements of
the code.
Finally, we~have improved and added to the figures and diagrams.

@@

A~full hidden Markov model also has distinguished ``begin'' and ``end''
states; begin and end states are like match states, but they
correspond to no amino acids and contain no emission probabilities.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, 
the begin and end states are known by their positions:
in a $\alignwidth$-column model, the begin state is the only state in
column~0 and the end state is the only state in column~$\alignwidth$.
Every other state is referred to by its column~$j$
and by one of these labels:\label{code:state-label}
\smallverbatiminput{statelabel}
We~group states that share a column~$j$; these states are referred
to collectively as ``node~$j$,'' as shown in \figref{plan7}.
This model
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.
\remark{We need to figure out if the begin state is really node 0}

@@

All scores computed by \mrfy\ are sums of these transition and
emission scores.
Because these sums can be usefully attached to many types of values,
we~have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \mono{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
%% \remark{Lazy in the first arg is about 1\%~slower}
The \texttt{/+/} function adds to the score without touching the
 container.
Function \texttt{fmap} is also defined; it~applies a function to a
 container's contents. 
%
Finally, we~made \texttt{Scored} an instance of~\texttt{Ord}.
Containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

@@

\begin{figure}%%%[h!]
\centerline{\loadpdf[]{speedup}}
\medskip
\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
\figlabel{speedup}
\end{figure}

@@

\seclabel{better-viterbi}
\seclabel{improving-viterbi}

@@

The code example for \verb+vee'+ on page~\pageref{code:vee-prime}
shows the computation of the general case of~$V^{\prime M}_j(i)$ from 
\figref{viterbi-transformed}.
But the full story is not nearly as nice at one could hope for:
in~the code that accompanies our 2012 ICFP paper,
\verb+vee'+ has 25~cases overall.
Many of these cases exist because we~did not have enough insight into
Haskell, and we made some decisions that might have been more
appropriate to a FORTRAN program:
\begin{itemize}
\item
Recursive computations were done on smaller array indices, not smaller
models.
\item
Array indices 1, 0, and $-1$ were all used to indicate special
conditions.
\item
Senseless combinations of array indices and states were handled by
assigning each such combination the largest representable score, which
corresponds to the smallest representable probability.
\end{itemize}
Before trying to speed up the code, we~simplified~it.
%
%
Our first step was to start thinking like Haskell
programmers and use types:
\begin{itemize}
\item
We~refactored the single, flat \texttt{Node} type from our ICFP paper
into the three state types you see on page~\pageref{code:model3-node}.
\item 
We introduced a \texttt{Model} type that consists of a special Begin
node containing a begin and an insertion state, an~unbounded sequence
of ordinary \texttt{Nodes}, and an end state.
And we represented the sequence as a list, not an array.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only match, insertion, and deletion states, so we removed the
\texttt{StateLabel} value constructors corresponding to begin and end
states, leaving only the three constructors shown on
page~\pageref{code:state-label}. 
(We~moved the begin and end labels to another type, which is used only
to prettyprint models.) 
\end{itemize}
The immediate effect of these changes was to reduce case analysis on
two dimensions:
instead of five state labels, we got three;
and instead of three special array indices, we got just one special
list (the empty list).

@@

The version of \verb+vee'+ shown in  \figref{viterbi-hov4} is abstract
in another way: the type of the result is an unknown type~\texttt{a}.
Type~\texttt{a} represents a tree-node in an abstract ``cost~tree;''
we~instantiate it as a score, as a \mono{Scored~[StateLabel]},
and as an explicit tree used to visualize and debug the computation.
Values of type~\texttt{a} are built using three functions that are
passed as parameters:
\smallverbatiminput{hosigs}
The \texttt{internal} function must treat an empty list of edges as an
unreachable tree-node; for~example, if \texttt{a} is instantiated
to \texttt{Score}, then \mono{internal~[]} must produce an infinite
score (corresponding to zero probability).

\figref{viterbi-hov4} is the first example in this paper that
shows \emph{all} the cases for~\verb+vee'+.
We~have reduced the original 25~cases to just~4.
The~general case, which was shown in other examples, appears last.
The three special cases above it handle the three states that can
follow the begin state.
In~our legacy file format, and also in our code, the begin state is
identified with the match state in node~$j=0$.
But although it is
identified with a match state, the begin state does not behave like a
match state: it~does not emit a residue.
Therefore the three states that can follow it can be entered only when the
count of residues emitted is~$i=0$.

Using the code in \figref{viterbi-hov4}, plus the auxiliary functions
shown above, the implementation of Viterbi's algorithm is just
25~lines of code, of which 2~lines are type signatures.
That's four times smaller than our original code.
But~while we've gained something, we've also lost something---it's not
at all obvious that the code in \figref{viterbi-hov4}  implements
Viterbi's equations.

Although the~new code is about 3\%~slower than our original code, 
it~turned out to a good starting point for experiments.
Of~the myriad variations we tried, 
the best involved two transformations:
\begin{itemize}
\item
We~abstracted the call to \texttt{internal} over the state~$\hat s$,
the list of preceding states, and functions that
compute \texttt{pi}~and~\texttt{pj}.
\item
We~then added a fourth case to \verb+vee'+, which handles the match
state in node~1 when the residue count is nonzero.
\end{itemize}
For reasons we don't understand, GHC specializes the resulting code
very effectively, leading to 30\%~faster performance on our
UNCHARACTERIZED MEDIUM benchmark.
The~result is 26\%~faster than our original code.

This change laid the groundwork for more experiments.
The most interesting of these experiments was to follow the advice of
\citet{tibell:high-performance} and use GHC's UNPACK pragma on scalar
values---in our case, scores.
Using the UNPACK pragma made our original code 7\%~faster but our new
code 9\%~slower.

We~then turned to GHC's profiler, running Viterbi's algorithm on a
model of 343~nodes and a query of 592~residues.
We~expect the memo table to have about 600,000 entries,
and the profiler says that the program allocates 531MB, or 7.85KB
per~entry.
This seems excessive.
\begin{itemize}
\item
One culprit is the function \verb+vee''+, which is charged with
allocating about 66MB.
To~estimate the amount of memory consumed by
an entry, we~imagine that representing a boxed
score on the heap might require three 32-bit words.
A~thunk involving three parameters might be a bit like a four-element
list, taking perhaps 12~words for the cons cells and another 8~words
for the values?
And then the pointer in the memo table itself would occupy a word.
So~24~words per entry times 4~bytes per word would be 96B per
entry for a total of about 59MB?
Because the actual memo table involves arrays of arrays of arrays,
we~won't quibble about a discrepancy of~10\%.
\item
The computation of \texttt{score} (the sum of transition 
and emission scores) is charged with allocating 45MB total, or about
4.5~words for each transition in the computation tree.
We~have no idea why this combination is allocating, and various
attempts to make the sum strict (e.g., by using a bang pattern in 
\mono{let~!score = }\ldots) either have no effect or actually make
performance worse.
\item
There is also some significant allocation in the parser, but the
elephant in the room is the function \verb+vee'+ itself, which
allocates 258MB, which is half the memory allocated by the entire
program.
We~have no idea what is happening.
A~graphical heap profile shows a large amount of memory initially
charged to \verb+vee''+ (the memo table), which then shrinks
gradually, presumably as large thunks are replaced with smaller
scores.
The~amount of memory charged to \verb+vee'+
increases roughly linearly with time, 
so~perhaps we have a space leak, but even given the various profiling
tools available, we~are not competent to find~it.
\end{itemize}
In~principle, we~could have diagnosed our problems by reading 
GHC Core output from the compiler.
We~found guidance in Don Stewart's invaluable blog posts,
and we also used his \texttt{ghc-core} tool.
But it wasn't enough.
Even though one of us has worked on several optimizing compilers, 
we~just don't have enough experience to use GHC Core to diagnose
allocation problems in a 25-line function.

%%\begin{verbatim}
%%### 2013-05-27 01:23   Running med(vpasses=1): icfp-branch mrfy nr-jfp
%%icfp-branch faab8dc       54.07s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
%%mrfy d5604bc*             44.90s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
%%nr-jfp 00ef54e*           47.59s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
%%\end{verbatim}

%%\smallverbatiminput{hov-prevs}

We~succeeded in making Viterbi's algorithm a lot simpler and somewhat
faster, but our original goal was to make it competetive with~C.
The measurements using SMURF are depressing, and the SMURF code is
even more depressing, so we reimplement Viterbi's algorithm from
scratch in a clean C~program.
To~avoid writing a parser, we extend \mrfy\ with code that reads a
model and a query sequence and emits a C~file containing that model
and query as initialized data.

If~we look not at the whole package but just at Viterbi's algorithm,
the~performance story is even more grim:
we~wrote a custom C~program that computes a score using Viterbi's
algorithm and does nothing else, and it runs about a hundred times
faster than \mrfy's implementation.
Ironically, we've~had a lot of trouble getting the code right, but it
allocates only one double-precision floating-point number for each
entry in the memo table; Haskell allocates about 30~times more.

It's depressing to think that a language with a mature, optimizing,
native-code compiler might not be usable for performance-critical
work, but the difficulty of diagnosing allocation problems is a real
obstacle.
We~were lucky that we could get new results (CITE NOAH'S LATEST PAPER
HERE?) even though our core algorithm is 100~times slower than the
same algorithm written in~C.
You~might not be so lucky.

\subsection{Debugging and testing: awkward, then  less awkward}

@@

Our experience writing code and trying new ideas was excellent, as~was
the ease of parallelizing \mrfy.
Higher-order functions, memoization, laziness, and parallel strategies
really worked.
But~as discussed above, profiling didn't really work for~us.
And we also encountered other obstacles that prevented functional
programming from working as well as we would have liked.
The~most significant such obstacles were in debugging and testing.

We had a hard time diagnosing run-time errors.
We~expected some run-time errors;
our~group's legacy file
format is poorly documented and hard to deal with.
(When beta
strands overlap and are doubly paired, even the invariants of the
format are unclear.)
But~using Haskell, we~found the errors hard to diagnose.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~didn't know about the backtrace feature of GHC's profiler,
and even after we learned about it, it didn't help:
the profiler can be used only on very small test cases,
which didn't always trigger the errors.
This same limitation affected GHCi's debugger; 
in~GHCi, our \texttt{vee'} function
is too slow to be runnable on nontrivial
inputs.
Moreover,  GHCi's debugger can set breakpoints only at top-level functions
or at specific column/line positions, which made debugging the memoized
\texttt{vee''} function impractical.
In~August~2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and when GHC~7.4 was released in early~2012, it~included support for
stack traces. 
We~have used the stack-trace feature a couple of times, and it helps.

@@

Only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles that prevented the
junior people from using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
(For programmers accustomed to object-oriented dispatch
on \emph{arguments}, it is hard to grasp how Haskell's type-class
system can find the proper version
of \texttt{arbitrary} using only the type of a \emph{value}.)
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did 
use QuickCheck, the counterexamples were too large to be
informative. 
(To~be fair to ourselves, the \texttt{shrink} method is barely
mentioned in the original 
paper on QuickCheck.
It~shows up under another name, as~an improvement made by Andy Gill.)
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only when the whole team pitched in to
meet deadlines for ICFP~2012.

After ICFP, 
QuickCheck played a central role in further developments.
Most important was the testing of the variations on Viterbi's
algorithm described in \secref{better-viterbi}.
We~used QuickCheck not only to shrink sequences of nodes and of amino
acids, but also to shrink the number of nonzero transmission and
emission probabilities in the models.
We~were able to confirm that each new implementation computed the same
paths with the same score as our original implementation.
Almost all counterexamples were shrunk to include just two nodes 
(the smallest model supported by our original implementation)
and two or three amino acids.
Moreover, the models contained just a handful of nonzero
probabilities.
QuickCheck played an invaluable role in helping us get edge cases
right, and in confirming that most of the cases in the original code
could be eliminated.

@@

If~you are a computational biologist and you are interested in
functional programming, you don't need extensive preparation
to be productive in Haskell.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell 
for medium-sized projects,
but he has played a secondary role, primarily refactoring and testing
others' code.

@@

Using GHC versions through~7.2, 
we~had quite some difficulty profiling.
These difficulties were technical in nature, and they have been
completely mitigated by the new profiling tools released 
with
GHC~7.4.
The~most useful new feature is the automatic annotation of ``cost
centers'' \cite{sansom-pj}. 
Some technical difficulties remain, particularly with installation:
the only reliable way we have found to enable compilation with
profiling is to destroy all existing compiled libraries, 
update our Cabal configuration to ask for libraries to be compiled
with profiling, and reinstall all needed Cabal packages.
As~of early 2013, the process of reinstallation, although not
difficult, is time-consuming and tedious.

@@

Ideally we would use QuickCheck to find bugs, but as we mention in
\secref{awkward-quickcheck}, there are obstacles.
We~have now overcome these obstacles, but
we~sorely regret not doing so earlier.
In~light of our experience, we~have instituted a new programming practice:
whenever we introduce a new data type,
we~write the instance of
\texttt{Arbitrary} right away, while
relevant invariants are still fresh in memory.
When invariants are not enforced by Haskell's static type system,
we~will write them as Haskell predicates.
We can then immediately run QuickCheck on each predicate, to verify
that our Arbitrary instance agrees with the predicate.
For each predicate~\texttt{p} we can also check
\mono{fmap (all p . shrink) arbitrary}.
We~cannot overemphasize the value of QuickCheck!

@@

If~you want to use Haskell in your research, we~believe that you must
have enough experience with functional programming that you can build
\emph{all} the code you need, not only the code that is easy to write
in a functional language.
Implementing Viterbi's equations in Haskell was pure joy.
And boiling down the code to just four cases, with the help of
QuickCheck, was also tremendous~fun.
Writing an iterative search in purely functional style was~easy.
Transforming data in the HMMER file format, \emph{without}
using mutable state the way the C++ code does, was difficult.

On~reviewing our work, we~observed that we were a little too ready to
use types and data structures better suited to C++ programming than to
Haskell programming.
Eventually we learned to step slowly away from the code and to seek
out opportunities for abstraction and simplification.
The~abstract, simple versions---supported by QuickCheck tests---are a
lot more likely to be useful to the next generation of graduate students.

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
John Tibell taught us the \texttt{UNPACK} pragma.
For~support and discussions 
we also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.

@@

%\section{THIS SECTION WILL BE DELETED.  SAUVE QUI PEUT!}

%\input{jfpnotes}
